Experience in Modeling a Microcontroller Instruction Set Using B

### 2012-01-22 papers/ArticleZ80/source/paper.tex

\institute{ Federal Institute of Education, Science and Technology of Rio Grande do Norte, Natal RN 59015-000, Brazil
           \and 
		    Federal University of Rio Grande do Norte, Natal RN 59078-970, Brazil}

@@

The rigorous development of an embedded software in a microcontroller requires a
specification of its instruction set and the software requirements specification.
The actual industrial practice is to document the instruction set informally
using natural language or with added semi-formal notation. This practice leaves
space for errors and inconsistencies \cite{springerlink:Yuan2011}. This paper
presents specifically a B formal model of the Z80 instruction set. This model 
can be  used to document, simulate, test and in a model-driven approach to
design and develop software from the functional specification level down to
assembly.
% Using this model-driven approach and the instruction set model, a small case
% study to analyse a petroleum production test system in oil field was developed.
%  \paragraph[Keywords]{ \textbf{Keywords}: Embedded Software, Simulation, B
% method and Verification%}
\end{abstract}

@@

This work presents a formal model of the instruction set of the Z80
microcontroller~\cite{Z80_manual}. This microcontroller was chosen
because of several factors: it contains essential and common concepts
of microcontrollers and even microprocessors; it has extensive
documentation available; and it has been widely used.  A first,
partial, Z80 model presented in \cite{Valerio_SBMF09} was fully
verified, but it uses constructs that make it impossible to animate.

The new model supports animation in ProB\cite{proB}.  Essentially, the
new model\footnote{The interested reader in more details is invited to
  visit our repository at: http://code.google.com/p/b2asm/.}  no
longer contains definitions derived from infinite sets. Using B
constructs for modularity, auxiliary libraries of basic modules were
developed as part of the construction of microcontroller model. Such
library has many definitions of concepts common in microcontrollers.
Besides the new Z80 model, a petroleum production test system was
developed using Z80 assembly language to analyse the code simulation
and verification process.

Providing the formal model of an instruction set yields many
benefits. The animation of Z80 model allows to simulate the execution
assembly programs including support for interrupts and input and
output instructions. Other possible uses of a formal model of a
microcontroller instruction set include documentation, the
construction of simulators, and can be possibly the starting point of
a verification effort for the actual implementation of a Z80
design. Moreover, the model of the instruction set could be
instrumented with non-functional aspects, such as the number of cycles
it takes to execute an instruction, to prove lower and upper bounds on
the execution time of a routine.  Two aspects of these formalization
are of particular importance to us: provide a more solid documentation
artifact for microcontrollers, build a reference model for a formal
compilation approach in B.

\subsection{Problems in formalizing instruction set}
% Compact version

Many times, each instruction of microcontrollers is shown in its official manual
using textual description, math description, examples, encoding, number of cycles
and other information. Sometimes, the instruction descriptions have
an own notation, some informations are organized in different sections
of document and the textual description are not standardized. 
The manuals also have semi-formal and informal descriptions, these descriptions
may leave errors, inconsistency and ambiguity.  
For example in official manual of the microcontroller Z80 \cite{Z80_manual}, if the users \marginpar{VM: Adicionei um exemplo real e uma referência  que cita vários problemas identificados no Manual do Z80.} 
want to know the action of an instruction and its effects on the flag register then he needs
to read an informal textual description, that  sometimes can be placed in different pages. 
Furthermore, the official manual of the microcontroller Z80 has also several
problems that were identified over time~\cite{UndocumentedZ80}.

A solution to avoid ambiguity and inconsistencies is to specify
formally the assembly instruction set, this creates a pattern of
representation and validates properties of instructions. The formal
model also restricts the definitions to correct typing and uses only
well-defined expressions. Furthermore, the developer can use with B
method different levels of abstraction and add more specific details
in refined models.

@@

This section presents the verification of the aforementioned modules
using AtelierB provers.  By default, AtelierB~\cite{atelierB} does
not generate proof obligations of the type of functions defined in the
clause $PROPERTIES$ of a machine. For example, the definition $inc \in
\{0,1\} \fun \{0,1\} \land inc = \lambda(x).( x \in \{0,1\} | x + 1)$,
does not create a proof obligation to check the consistency of the
definition. However, this definition is inconsistent since $inc(1)=2$
and $ 2 \not\in \{0,1\}$.

To avoid such inconsistencies, we state separately the required properties in the
\textit{ASSERTIONS\/} clause, that cause corresponding proof obligations to be
generated and verified. However, we have not been able to build the corresponding
proofs using the base of rules included in the free distribution of AtelierB. So
we formalized new proof rules to check these properties in AtelierB and used them
in the interactive prover of AtelierB.

The hardware library has many conversion functions with
arithmetic,\marginpar{Não seria interessante ter colocado
  esta figura mais cedo? Na seção 2.5, por exemplo? E no texto da
  seção 2.6 poderia enumerar os poucos casos verificados sem regras
  adicionais.}  for example, Bit\_vector to Natural, Natural to
Bit\_vector, Integer to Bit\_vector and Bit\_vector to Integer. These
functions are organized in modules and each module contains auxiliary
functions for a data type. The modules are represented by gray
rectangles in Figure \ref{DiagramTypesRules}.

%and they have the mean: \textit{BYTE} is a bit vector with eight elements enumerated
%starting from one; \textit{UCHAR} is a range between 0 and 255; \textit{SCHAR} is a range between -128 and 127; 
%\textit{BV16} is a bit vector with sixteen elements enumerated starting from one; \textit{USHORT} is a range 
%between 0 and 65535; and \textit{SSHORT} is a range between -32768 and 32767.

@@

%\textbf{We needed to verify the type of these conversion
%  functions}.\marginpar{Não precisa repetir
%  isto.} % There are twenty rules associated this properties
The conversion functions need to satisfy two properties: to be
bijective and equal to the reverse of the dual conversion
function. The three simplest proof obligations were verified
automatically with AtelierB: (\it uchar\_schar\rm \ $\in$
\textit{UCHAR} $\bij$ \textit{SCHAR}), (\it schar\_uchar\rm \ $\in$
\textit{SCHAR} $\bij$ \textit{UCHAR}) and (\it byte\_bv16\rm \ $=$ \it
bv16\_byte\rm $^{-1}$).
%\begin{enumerate}
%
%\item (\it uchar\_schar\rm \ $\in$ \textit{UCHAR} $\bij$ \textit{SCHAR})
%\item (\it schar\_uchar\rm \ $\in$ \textit{SCHAR} $\bij$ \textit{UCHAR})
%\item (\it byte\_bv16\rm \ $=$ \it bv16\_byte\rm $^{-1}$) 
%
%\end{enumerate}
For the remaining proof obligations, we again had to create new proof
rules. There are three kinds of proof rules, that we illustrate by giving
one example each:

@@

The Z80 formal model provides many benefits, because of the verification of
generated proof obligations guarantees: the correct use of data types, the developed
security properties  and the well-defined of all the expressions. %\footnote{An expression is called ``well-defined'' (or unambiguous) if its definition assigns it a unique interpretation or value.}.
Furthermore, the designer has a big flexibility to create new and specific security properties, this is very useful
to adjust the verification in accordance with the requirements. Moreover, this
example of model could replace or improve the used documentation for users and
assembly programmers. Z80 model is also useful to develop verified software 
up to the assembly level. There are many benefits when a formal model is
developed, but to verify the model is not an easy task, the first version of Z80 model have already verified completely,but the new Z80 model with support to animation in ProB has not been verified completely,
because the changes in Z80 model are recent and the verification time is long.

A case study using the first version of Z80 model was also created and verified
since the first abstract B model up to B assembly model. This case study is an
object of the pilot project developed to analyse a petroleum production test
system in each oil field. Its modeling was developed in according to \cite{LAUT_SERGIO}
and its B assembly model can be executed manually in ProB. The modeling of
instruction set and the B assembly model allow to simulate the execution of microcontroller
and analyse several important properties.

@@

There are in the literature several similar approaches to model hardware and instruction set using B.
The paper \cite{springerlink:Yuan2011} reports a method to specify, design and construct sound and complete instruction set model by stepwise refinement and formal proof using the formal method Event-B. It also discuss about desirable properties of instruction set model. However, our work \cite{Valerio_SBMF09} and \cite{Subotic2010} use the B method, that seems more appropriated to software development, because it has implementable language defined, 
called B0, and tools to convert the models to a programming language.
In \cite{Marc20113} is present an related experience on using B in the design of secure micro-controllers. It tries to show the feasibility of such a technique for high confidence trustful devices. The work \cite{Subotic2010} describes a similar approach with MIPS CPU architecture and it shows how the CPU is formally specified and implemented using the B method.

@@

% Um pequeno resumo do trabalho
This work has shown an approach to the formal modeling of the instruction set of
microcontroller using the B method. During the construction of this model, some
problems were found in the official reference for Z80
microcontroller~\cite{Z80_manual}. The designer fixed the found problems in
documentation and also developed a case study: a verified embedded software. This
case study was interesting to analyse the technique of formal verification up to
assembly level.

% TODO: Adaptar para apenas fundamentar melhor o trabalho já desenvolvido. Citar o artigo SBMF e a dissertação. 
% Os meus trabalhos já desenvolvidos/ resultados atuais.
The following works quoted are directly related to the objective this paper.
The first work \cite{DantasSemish2008} shows the developed metodology to verify
software up to assembly level using B. A second work\cite{Dantas_SBMF08}
presents more details about the verification approach and a small 
software that was verified up to assembly level in three different platforms. 
The last work \cite{Valerio_SBMF09} shows a general view of the first version of Z80
model and techniques used in verification process.
%Link com o parágrafo anterior 
These works created an important step to software verification up to assembly
language, they show the current difficulties and suggest improvements in the
B tools.

@@

Future works comprise the development of software with the B method from
functional specification to assembly level, using the Z80 model presented in this
work. The mechanic compilation from B algorithmic constructs to assembly platform
is also envisioned. Finally, another important activity is to develop a formal
model of a platform used in LLVM Compiler~\cite{DBLP:conf/cgo/LattnerA04}.

>>>>

\institute{Federal Institute of Education, Science and Technology of Rio Grande do Norte, Natal, RN, Brazil
  \and 
  Federal University of Rio Grande do Norte, Natal, RN, Brazil}

@@

  This paper presents a formal model, developed in B, of the Z80
  instruction set. The formal model includes different libraries that
  specify entities commonly used in microcontrolers and
  microprocessors. Not only this model can be used for documentation
  purposes, but also to simulate and test the assembly code. It is also
  a fundamental component to extend the reach of the B method from the
  specification of the functional requirements of a system to an assembly-level
  implementation thereof.
% Using this model-driven approach and the instruction set model, a small case
% study to analyse a petroleum production test system in oil field was developed.
%  \paragraph[Keywords]{ \textbf{Keywords}: Embedded Software, Simulation, B
% method and Verification%}
\end{abstract}

@@

This work presents a formal model of the instruction set of the Z80
microcontroller~\cite{Z80_manual}. This microcontroller was chosen
because of several factors: it contains essential and common concepts
of microcontrollers and even microprocessors; it has extensive
available documentation; and it has been widely used.  A first,
partial, Z80 model presented in \cite{Valerio_SBMF09} was fully
verified, but it used constructs that made it impossible to animate.

The new model supports animation in ProB\cite{proB}.  Essentially, the
new model\footnote{The interested reader in more details is invited to
  visit our repository at: http://code.google.com/p/b2asm/.}  no
longer contains definitions derived from infinite sets. Using B
constructs for modularity, auxiliary libraries of basic modules were
developed as part of the construction of microcontroller model. Such
library has many definitions of concepts common in microcontrollers.
Besides the new Z80 model, a petroleum production test system was
developed using Z80 assembly language to analyse the code simulation
and the verification process.

The formal model of an instruction set has several
applications. Animation allows to simulate the execution of assembly
programs including support for interrupts, input and output
instructions. Other possible uses include documentation, the
construction of simulators, and can be possibly the starting point of
a verification effort for the actual implementation of a Z80
design. Moreover, the model of the instruction set could be
instrumented with non-functional aspects, such as the number of cycles
it takes to execute an instruction, to prove lower and upper bounds on
the execution time of a routine.  Two aspects of such formalization
are of particular importance to us: provide a more solid documentation
artifact for microcontrollers and build a reference model for a formal
compilation approach in B.

It is often the case that the official documentation of an instruction
uses textual description, mathematical formulas, examples,
etc. Sometimes, the instruction descriptions have a specific notation,
some informations are organized in different sections of document and
the textual descriptions are not standardized.  The manuals also have
semi-formal and informal descriptions, these descriptions may leave
errors, inconsistency and ambiguity.  For example, in the official
manual of the microcontroller Z80 \cite{Z80_manual}, the semantics of
an instruction (in particular its effects on the flag register) are
described informally, and is this description is scattered in
different pages.  Furthermore, the official manual of the
microcontroller Z80 has also several problems that were identified
over time~\cite{UndocumentedZ80}.

A solution to avoid ambiguity and inconsistencies is to specify
formally the assembly instruction set, making it possible to prove
properties of the instructions. The formal model also restricts the
definitions to correct typing and uses only well-defined
expressions. Furthermore, the developer may use the refinement
capabilities of the B method to specify at different levels of
abstraction and add more specific details or useful restrictions in
refined models.

@@

This section presents the verification of the aforementioned modules
using AtelierB provers.  By default, AtelierB~\cite{atelierB} does not
generate proof obligations to check the type of functions defined in
the $PROPERTIES$ clause of a machine. For example, the definition $inc
\in \{0,1\} \fun \{0,1\} \land inc = \lambda(x).( x \in \{0,1\} | x +
1)$, does not create a proof obligation to check the consistency of
the definition. However, this definition is inconsistent since
$inc(1)=2$ and $ 2 \not\in \{0,1\}$.

To avoid such inconsistencies, we state separately the required
properties in the \textit{ASSERTIONS\/} clause: this causes
corresponding proof obligations to be generated and verified. However,
we have not been able to build the corresponding proofs using the base
of rules included in the free distribution of AtelierB. So we
formalized new proof rules to check these properties in AtelierB and
used them in the interactive prover of AtelierB.

The hardware library has many conversion functions with
arithmetic,  for example, Bit\_vector to Natural, Natural to
Bit\_vector, Integer to Bit\_vector and Bit\_vector to Integer. These
functions are organized in modules and each module contains auxiliary
functions for a data type. The modules are represented by gray
rectangles in Figure \ref{DiagramTypesRules}.

@@

The conversion functions need to satisfy two properties: bijectivity
and equality to the reverse of the dual conversion function. Three
such proof obligations were verified automatically with AtelierB: (\it
uchar\_schar\rm \ $\in$ \textit{UCHAR} $\bij$ \textit{SCHAR}), (\it
schar\_uchar\rm \ $\in$ \textit{SCHAR} $\bij$ \textit{UCHAR}) and (\it
byte\_bv16\rm \ $=$ \it bv16\_byte\rm $^{-1}$).
%\begin{enumerate}
%
%\item (\it uchar\_schar\rm \ $\in$ \textit{UCHAR} $\bij$ \textit{SCHAR})
%\item (\it schar\_uchar\rm \ $\in$ \textit{SCHAR} $\bij$ \textit{UCHAR})
%\item (\it byte\_bv16\rm \ $=$ \it bv16\_byte\rm $^{-1}$) 
%
%\end{enumerate}
For the remaining proof obligations, we again had to create new proof
rules. There are three kinds of proof rules, that we illustrate by giving
one example each:

@@

The Z80 formal model provides many benefits, because of the verification of
generated proof obligations guarantees: the correct use of data types, the developed
security properties  and the well-defined of all the expressions. %\footnote{An expression is called ``well-defined'' (or unambiguous) if its definition assigns it a unique interpretation or value.}.
Furthermore, the designer has a big flexibility to create new and specific security properties, this is very useful
to adjust the verification in accordance with the requirements. Moreover, this
example of model could replace or improve the used documentation for users and
assembly programmers. Z80 model is also useful to develop verified software 
up to the assembly level. %There are many benefits when a formal model is
%developed, but to verify the model is not an easy task, the first version of Z80 model have already verified %completely,but the new Z80 model with support to animation in ProB has not been verified completely,
%because the changes in Z80 model are recent and the verification time is long.

A case study using the first version of Z80 model was also created and verified
since the first abstract B model up to B assembly model. This case study is an
object from the pilot project developed to analyse a petroleum production test
system in each oil field. Its modeling was developed according to \cite{LAUT_SERGIO}
and its B assembly model can be executed manually in ProB. The modeling of
instruction set and the B assembly model allow to simulate the execution of microcontroller
and analyse several important properties.

@@

There exists several approaches to model hardware and instruction set
using B a.  The paper \cite{springerlink:Yuan2011} reports a method to
specify, design and construct sound and complete instruction set model
by stepwise refinement and formal proof using Event-B. It also
discusses desirable properties of the instruction set model. However,
our work \cite{Valerio_SBMF09} and \cite{Subotic2010} use the B
method, that seems quite appropriated to software development, because
it has an implementable language defined, called B0, a model-driven
approach \cite{Dantas_SBMF08} to develop software from the functional
specification level down to assembly, and tools to convert the models
to a programming language.

A related experience on using B in the design of secure
micro-controllers is present in \cite{Marc20113}. It tries to show the
feasibility of such a technique for high confidence trustful
devices. The work \cite{Subotic2010} describes a similar approach with
MIPS CPU architecture and it shows how the CPU is formally specified
and implemented using the B method.

@@

% Um pequeno resumo do trabalho
This work has shown an approach to the formal modeling of the
instruction set of microcontroller using the B method.  During the
construction of this model, some problems were found in the official
reference for Z80 microcontroller~\cite{Z80_manual}. We have consulted
different sources \cite{Simulator_z80,UndocumentedZ80,Z80_manual} to
avoid such problems in modeling and have developed a case study: a
verified embedded software.  This case study was interesting to as a
feasibility analysis of the approach to derive formally assembly code
from a specification of functional requirements in the B method.

% TODO: Adaptar para apenas fundamentar melhor o trabalho já desenvolvido. Citar o artigo SBMF e a dissertação. 
% Os meus trabalhos já desenvolvidos/ resultados atuais.
The following works are directly related to the objective this paper.
The approach to develop verified software down to the assembly level
using B was described in~\cite{DantasSemish2008}. A first case study
for this approach was reported in~\cite{Dantas_SBMF08}, presenting
more details as well as a small software that developed up to assembly
level in three different platforms. A general view of a previous used
in verification process were presented in~\cite{Valerio_SBMF09}. The
model presented in the current paper added support for animation in
ProB and the representation of external actions (interruptions).
%Link com o parágrafo anterior 
This set of papers present an important step to software verification
up to assembly language, showing the current difficulties and
suggesting improvements in tools supporting the B method.

@@

In the future, we plan to improve the support for the development of
software with the B method from functional specification to assembly
level, using the Z80 model presented in this work. For instance, the
mechanic compilation from B algorithmic constructs to assembly
platform is also envisioned. Finally, another important activity is to
develop a formal model of a platform used in LLVM
Compiler~\cite{DBLP:conf/cgo/LattnerA04}, which would enable us to
integrate different compilation techniques into our B-based framework.
